---
title: "Do It Again, R!"
author: "Jeff Newmiller"
date: "7 May 2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( echo = TRUE )
```

## Outline

- For Loops
    + Where is my output?
    + Keeping results
- Vectors of inputs and vectors of outputs
    + Vectorized functions, not "Vectorize my function"
    + Appending columns
    + Plotting results: ggplot2
- Expanding options
    + Cartesian join: expand.grid
- Compressing options: Subset, subset, subset... Grouping!
  + Truncation
  + "cut"-ing
  + switch
  + forcats
- "ave" and "aggregate"... or "group_by/mutate" and "group_by/summarise"
- Lists

## Packages Used

```{r loadpackages,echo=FALSE,warning=FALSE}
suppressPackageStartupMessages( library( dplyr ) )
suppressPackageStartupMessages( library( ggplot2 ) )
```


```{r}
library(dplyr)
library(ggplot2)
```


## Sample Problem

Given a relationship between $x$ and $y$:

$$
y = a x^2 + b x + c
$$

we can find which values of $x$ cause $y=0$ using the quadratic formula:

$$
x = \frac{-b \pm \sqrt{b^2-4ac}}{2a}
$$

## Sample Solution

For now, we will focus on the most positive answer:

```{r quad1}
A <- 1
B <- -2
C <- -3
DF1 <- tibble( x = seq( -3, 5, 0.1 )
             , y = A*x^2 + B*x + C
             )
DF1b <- tibble( x = 3, y = 0 )
ans <- ( -B + sqrt( B^2 - 4*A*C ) ) / ( 2 * A )
ans
```

## Solution Plot

```{r,echo=FALSE}
ggplot( DF1, aes( x = x, y = y ) ) +
  geom_line() +
  geom_point( data = DF1b, size = 2, colour = "blue" ) +
  theme_minimal()
```


## Do it more than once?

```{r readmyabc}
input <- read.csv( "data/MyABC1.csv" )
input
```


## For Loop - Printing

A common approach used in other languages is to print results as you go along:

```{r forprint, echo = TRUE}
for ( i in seq_along( input$A ) ) {
    A <- input$A[ i ]
    B <- input$B[ i ]
    C <- input$C[ i ]
    print( sprintf( "A = %8.4f, B = %8.4f, C = %8.4f, x = %8.4f"
                  , A, B, C
                  , ( -B + sqrt( B^2 - 4*A*C ) )
                    / ( 2 * A ) ) )
}
```

## For Loop - Concatenating

Another common recommendation is to concatenate:

```{r forconcat}
ans <- numeric( 0 )
for ( i in seq_along( input$A ) ) {
    A <- input$A[ i ]
    B <- input$B[ i ]
    C <- input$C[ i ]
    ans <- c( ans, ( -B + sqrt( B^2 - 4*A*C ) ) / ( 2 * A ) )
}
ans
```

## For Loop - Element Replacement

Concatenating can be _very_ inefficient if there are many elements of `ans` because there are $n-1$ shorter versions of `ans` created and discarded along the way.

```{r formodify}
ans <- numeric( nrow( input ) )
ans
for ( i in seq_along( input$A ) ) {
    A <- input$A[ i ] ; B <- input$B[ i ] ; C <- input$C[ i ]
    ans[ i ] <- ( -B + sqrt( B^2 - 4*A*C ) ) / ( 2 * A )
}
ans
```

## Combined Presentation

Having your results in a vector of their own is useful, but showing that vector next to the inputs is even more useful:

```{r}
cbind( input, tibble( ans = ans ) )
```

## Make a Solution Function

To make a function, write code that uses inputs mentioned in the parentheses, and the last expression in the function will be the answer given by the function:

```{r}
my_solution <- function( A, B, C ) {
  ( -B + sqrt( B^2 - 4*A*C ) ) / ( 2 * A )
}
my_solution( 1, -2, -3 ) # try it out
```

You can also draw on variables from the working environment where you define the function, but it is intentionally hard to make changes to such variables inside a function.

## Make a Many Solutions Function

If your solutions generally only worked for one value each for `A`, `B`, and `C` then you might make this function:

```{r}
my_many_solutions <- function( A, B, C, solution ) {
  ans <- numeric( length( A ) )
  for ( i in seq_along( A ) ) {
    ans[ i ] <- solution( A[ i ], B[ i ], C[ i ] )
  }
  ans
}
my_many_solutions( input$A, input$B, input$C, my_solution )
```

## But why Re-Invent the Wheel?

The `pmap_dbl` function takes a data frame and applies a function to each row. The function arguments have to be the same as the column names in the data frame, so we exclude the `Trial` column before giving it to `pmap_dbl`:

```{r}
library(purrr)
pmap_dbl( input[ , -1 ], my_solution )
```

## Vectorized Calculations

In this case, because R arithmetic operations are internally vectorized, the `my_solution` function can directly be given vectors instead of one value at a time, which is much faster:

```{r}
my_solution( input$A, input$B, input$C )
```

## Why Not Always Vectorize?

Some algorithms are hard to implement in vectorized fashion... you may reach your goal sooner if you use a slower but simpler algorithm that is imperfect but gets the job done: 

```{r}
my_slow_solution <- function( A, B, C ) {
  Y <- function( X, A, B, C ) { A * X^2 + B * X + C }
  # only one answer at a time
  uniroot( Y, c( 0, 1e6 ), A = A, B = B, C = C, tol = 1e-8 )$root
}
my_slow_solution( 1, -2, -3 )
```

but if you plan to solve many thousands or millions of problems, it may pay off to spend time making a complex algorithm vectorized. 

## Speed Review

The price of flexibility when using `pmap_dbl` is speed.

```{r}
library(microbenchmark)
ans <- microbenchmark( 
  my_many_slow = my_many_solutions( input$A, input$B, input$C, my_slow_solution )
, pmap_slow = pmap_dbl( input[ , -1 ], my_slow_solution )
, loop_single = my_many_solutions( input$A, input$B, input$C, my_solution )
, pmap_single = pmap_dbl( input[ , -1 ], my_solution )
, vectorized = my_solution( input$A, input$B, input$C )
)
knitr::kable( summary( ans ) )
```



